---
title: "Differentiation timing score"
description: "PCA-based scoring of differentiation timing from bulk RNA-seq time-course data."
site: bookdown::bookdown_site
documentclass: book
---

```{r meta-versions, include=FALSE}
tutorial_dependency_packages <- c(
  'AnnotationDbi',
  'bookdown',
  'circlize',
  'clusterProfiler',
  'ComplexHeatmap',
  'DESeq2',
  'edgeR',
  'ggplot2',
  'ggrepel',
  'knitr',
  'org.Hs.eg.db',
  'patchwork',
  'ragg',
  'scales',
  'systemfonts',
  'viridis'
)
tutorial_dependency_text <- paste(
  sprintf(
    '%s %s',
    tutorial_dependency_packages,
    vapply(
      tutorial_dependency_packages,
      function(package_name) as.character(utils::packageVersion(package_name)),
      character(1)
    )
  ),
  collapse = '; '
)
tutorial_bioc_version <- as.character(BiocManager::version())
```

<div class="tutorial-title-block">
<h1 class="title">Differentiation timing score</h1>
<div class="tutorial-subtitle">PCA-based scoring of differentiation timing from bulk RNA-seq time-course data</div>

<div class="tutorial-meta-block">
<div><span>Author:</span> Zoheb Khan</div>
<div><span>Created:</span> 2025-08-29</div>
<div><span>Modified:</span> 2026-08-08</div>
<div><span>Compiled:</span> `r format(Sys.Date(), '%Y-%m-%d')`</div>
<div><span>Environment:</span> R `r as.character(getRversion())`; Bioconductor `r tutorial_bioc_version`</div>
<div><span>Dependencies:</span> `r tutorial_dependency_text`</div>
<div><span>Source:</span> <a href="https://github.com/ZohebKhan1/pca-maturation-scoring">github.com/ZohebKhan1/pca-maturation-scoring</a></div>
</div>
</div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center'
)

tutorial_figure <- function(filename, alt_text, width = '96%') {
  cat(
    sprintf(
      '<img class="tutorial-figure" src="assets/figures/%s" alt="%s" style="width:%s;">',
      filename,
      alt_text,
      width
    )
  )
}
```

# Introduction

Bulk RNA-seq differentiation studies are typically sampled at discrete timepoints, even though individual samples may progress through differentiation at different rates. This tutorial develops a reference-trajectory method that uses gene expression to estimate each sample's continuous position along a known differentiation time course.

The method uses reference samples to define both the expression space and the trajectory through that space. It then assigns each sample a predicted reference time, a normalized differentiation score, and a distance from the fitted trajectory.

The workflow has four operations:

1. select temporal genes from the reference samples;
2. train a centered, unscaled PCA model on the reference-timepoint mean profiles;
3. join ordered day centroids into a piecewise-linear trajectory; and
4. project samples to their nearest position on that finite trajectory.

The worked analysis is maintained in `scripts/01_build_tutorial_objects.R`. The reusable public implementations are `functions/select_temporal_genes.R` and `functions/score_differentiation_timing.R`. `tutorial/tutorial.Rmd` is the authoritative tutorial source; `docs/` contains the generated GitHub Pages site.

## Reuse the scorer

The scorer assumes that temporal genes and normalized expression values have already been defined. It trains the reference PCA and day-centroid polyline, projects all supplied samples into that fixed space, and returns predicted time, a normalized differentiation score, and projection-distance diagnostics. It does not select genes, normalize counts, draw figures, or write files.

```{r download-scoring-function, eval=FALSE, echo=TRUE}
base_url <- paste0(
  'https://raw.githubusercontent.com/ZohebKhan1/',
  'pca-maturation-scoring/main/functions'
)

download.file(
  paste0(base_url, '/score_differentiation_timing.R'),
  'score_differentiation_timing.R'
)
source('score_differentiation_timing.R')
```

```{r scorer-input-table, echo=FALSE, results='asis'}
scorer_input_table <- data.frame(
  Parameter = paste0(
    '<code>',
    c(
      'expression_matrix',
      'metadata',
      'temporal_genes',
      'sample_id_col',
      'time_col',
      'reference_samples'
    ),
    '</code>'
  ),
  Default = c(
    'Required',
    'Required',
    'Required',
    '<code>"sample_id"</code>',
    '<code>"day_numeric"</code>',
    '<code>NULL</code>'
  ),
  Purpose = c(
    'Normalized genes-by-samples expression matrix with unique row and column IDs.',
    'One row per expression sample; sample IDs must match the matrix columns.',
    'Temporal-gene IDs selected independently of the samples being evaluated.',
    'Metadata column containing unique sample IDs.',
    'Metadata column containing numeric reference times.',
    'Sample IDs used to train the reference trajectory; all samples are used when omitted.'
  ),
  stringsAsFactors = FALSE
)

cat('<div class="table-scroll" role="region" aria-label="Scorer input parameters" tabindex="0">')
knitr::kable(
  scorer_input_table,
  format = 'html',
  escape = FALSE,
  table.attr = 'class="sample-summary-table parameter-table"',
  align = c('l', 'l', 'l'),
  caption = 'Inputs to score_differentiation_timing().'
)
cat('</div>')
```

The scorer automatically retains the smallest number of PCs explaining at least 99% of the between-timepoint variance. The output list contains `scores`, `pca_coordinates`, `pca_fit`, `pca_variance`, the selected `n_pcs`, `centroid_polyline`, retained `temporal_genes`, and the `reference_time_range`. Because projection is clipped to finite centroid-to-centroid segments, predicted times and normalized scores are bounded by the first and last reference timepoints. A score outside 0 to 1 is therefore not part of this implementation’s contract.

# Example data used in the workflow/tutorial

The worked example uses processed data derived from public [NCBI GEO accession GSE122380](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122380), a bulk RNA-seq time course of human induced pluripotent stem cell differentiation into cardiomyocytes reported by Strober, Elorbany, Rhodes and colleagues. The analysis includes 192 samples from 13 cell lines collected across differentiation days 1 through 15. These samples form the reference cohort used for temporal-gene selection, PCA training, day-centroid trajectory fitting, and internal validation.

The repository bundles aligned sample metadata, a raw gene-count matrix, and a variance-stabilized expression matrix. The active input contains no perturbation samples. These tracked RDS files are the reproducible starting point for the workflow; raw-read acquisition, sequencing and sample-level QC, alignment, quantification, upstream gene filtering, and construction of the VST matrix are outside the repository's scope. The tutorial therefore reproduces the analysis from processed inputs rather than from FASTQ files.

```{r sample-summary-table, echo=FALSE, results='asis'}
sample_summary <- stats::aggregate(
  sample_id ~ day_numeric,
  data = metadata,
  FUN = length
)
cell_line_summary <- stats::aggregate(
  cell_line ~ day_numeric,
  data = metadata,
  FUN = function(cell_lines) length(unique(cell_lines))
)
data_summary <- merge(sample_summary, cell_line_summary, by = 'day_numeric')
names(data_summary) <- c('day', 'Total samples', 'Unique cell lines')
data_summary <- data_summary[order(data_summary$day), ]

summary_table <- rbind(
  Samples = data_summary[['Total samples']],
  `Cell lines` = data_summary[['Unique cell lines']]
)
summary_table <- data.frame(
  Metric = rownames(summary_table),
  summary_table,
  row.names = NULL,
  check.names = FALSE
)
names(summary_table)[-1] <- paste0('D', data_summary$day)

cat('<div class="table-scroll" role="region" aria-label="Samples and cell lines by differentiation day" tabindex="0">')
knitr::kable(
  summary_table,
  format = 'html',
  table.attr = 'class="sample-summary-table"',
  align = c('l', rep('r', ncol(summary_table) - 1L)),
  caption = 'Samples and represented cell lines by differentiation day.'
)
cat('</div>')
```

The overview below shows global time-course structure and the corresponding day-collapsed expression similarity.

```{r reference-pca-and-correlation, echo=FALSE, results='asis'}
cat('<div id="fig-reference-overview" class="figure-anchor"></div>')
tutorial_figure(
  'GSE122380_reference_pca_and_day_correlation.png',
  'PCA of the reference samples and a day-level expression correlation heatmap.'
)
```

<div class="tutorial-figure-caption"><strong>Figure 1.</strong> Experimental structure of the GSE122380 time course. <strong>a.</strong> PCA using the top 10% most variable VST genes, colored by differentiation day. <strong>b.</strong> Pearson correlation between day-mean profiles from the same variable-gene set.</div>

# Scoring workflow

<ol class="workflow-flow" aria-label="Differentiation timing scoring workflow">
  <li>
    <div class="workflow-step-title">Select temporal genes</div>
    <div class="workflow-step-text">Apply expression, time-effect, and VST-range filters within the reference samples.</div>
  </li>
  <li>
    <div class="workflow-step-title">Train the reference PCA</div>
    <div class="workflow-step-text">Fit centered, unscaled PCA to the mean temporal-gene profile at each reference timepoint.</div>
  </li>
  <li>
    <div class="workflow-step-title">Construct the time trajectory</div>
    <div class="workflow-step-text">Connect the ordered reference-timepoint centroids with finite line segments.</div>
  </li>
  <li>
    <div class="workflow-step-title">Project samples</div>
    <div class="workflow-step-text">Return predicted time, normalized score, and distance from the reference trajectory.</div>
  </li>
</ol>

## Select temporal genes

Temporal-gene selection applies three filters in sequence. A gene is retained only if it satisfies every criterion.

The DESeq2 likelihood-ratio test always asks whether categorical time improves
the count model. Users specify only the metadata columns that should be adjusted
for, such as batch, donor, sex, or another known nuisance variable. The function
then constructs the two models below.

```{r lrt-design-table, echo=FALSE, results='asis'}
lrt_design_table <- data.frame(
  Model = c('Full', 'Reduced'),
  Formula = c(
    '<code>~ adjustment_covariates + factor(time_col)</code>',
    '<code>~ adjustment_covariates</code>'
  ),
  Purpose = c(
    'Models adjustment variables and the categorical time effect.',
    'Models the same adjustment variables without time.'
  ),
  stringsAsFactors = FALSE
)

cat('<div class="table-scroll" role="region" aria-label="General DESeq2 likelihood-ratio design" tabindex="0">')
knitr::kable(
  lrt_design_table,
  format = 'html',
  escape = FALSE,
  table.attr = 'class="sample-summary-table parameter-table design-table"',
  align = c('l', 'l', 'l'),
  caption = 'General DESeq2 likelihood-ratio design for temporal-gene selection.'
)
cat('</div>')
```

With no adjustment covariates, the reduced model is `~ 1`. DESeq2 calculates
\(\mathrm{LR}=2(\ell_{\mathrm{full}}-\ell_{\mathrm{reduced}})\), where
\(\ell\) is the maximized model log likelihood. A small adjusted \(p\)-value
therefore indicates that categorical time improves fit after accounting for the
selected covariates; the test does not assume a linear expression trend.

```{r selector-parameter-table, echo=FALSE, results='asis'}
selector_parameter_table <- data.frame(
  Parameter = paste0(
    '<code>',
    c(
      'raw_counts',
      'vst_expression',
      'metadata',
      'sample_id_col',
      'time_col',
      'adjustment_covariates',
      'reference_group_col',
      'reference_group_value',
      'expression_cpm_cutoff',
      'lrt_padj_cutoff',
      'vst_dynamic_range_cutoff'
    ),
    '</code>'
  ),
  Default = c(
    'Required',
    'Required',
    'Required',
    '<code>"sample_id"</code>',
    '<code>"day_numeric"</code>',
    '<code>NULL</code>',
    '<code>NULL</code>',
    '<code>NULL</code>',
    '<code>10</code>',
    '<code>1e-7</code>',
    '<code>0.6</code>'
  ),
  Purpose = c(
    'Raw integer counts used for TMM CPM and the DESeq2 LRT.',
    'VST genes-by-samples expression matrix aligned to the raw counts.',
    'One row per matrix sample with the requested design and reference columns.',
    'Metadata column containing unique sample IDs.',
    'Numeric time column; modeled categorically in the LRT.',
    'Metadata columns included in both models; use NULL for no adjustment.',
    'Optional metadata column identifying reference groups.',
    'Value in reference_group_col defining the samples used for gene selection.',
    'Minimum maximum day-mean TMM CPM.',
    'Maximum adjusted p-value for the categorical-time LRT.',
    'Minimum range across day-mean VST values.'
  ),
  stringsAsFactors = FALSE
)

cat('<div class="table-scroll" role="region" aria-label="Temporal gene selection parameters" tabindex="0">')
knitr::kable(
  selector_parameter_table,
  format = 'html',
  escape = FALSE,
  table.attr = 'class="sample-summary-table parameter-table"',
  align = c('l', 'l', 'l'),
  caption = 'Parameters for select_temporal_genes().'
)
cat('</div>')
```

```{r temporal-filter-code, eval=FALSE, echo=TRUE}
source('functions/select_temporal_genes.R')

temporal_selection <- select_temporal_genes(
  raw_counts = counts,
  vst_expression = vst,
  metadata = metadata,
  sample_id_col = 'sample_id',
  time_col = 'day_numeric',
  adjustment_covariates = 'cell_line' # GSE122380-specific
)
temporal_genes <- temporal_selection$temporal_genes
```

The three filtering thresholds default to the values used above: maximum
day-mean TMM CPM at least 10, LRT adjusted \(p\)-value below \(10^{-7}\), and
day-mean VST range at least 0.6. `adjustment_covariates` defines the metadata
columns included in both LRT models; use `NULL` for an unadjusted time test or a
character vector such as `c('cell_line', 'batch')` for multiple covariates. The
function adds categorical time to the full model automatically. Numeric
adjustment columns remain continuous; character and logical columns are treated
as factors. This worked dataset uses `cell_line`; another study might use
`adjustment_covariates = 'batch'` or `c('batch', 'donor')` instead.

To select temporal genes from one reference cohort, provide a metadata column
and value together. For example,

```r
reference_group_col = 'condition',
reference_group_value = 'control'
```

restricts CPM filtering, the LRT, and VST-range filtering to samples labeled
`control` in `metadata$condition`. Omitting both arguments uses all samples.
This pair controls which samples enter gene selection; it is not a DESeq2
coefficient reference level, which does not affect this omnibus LRT.

```{r lrt-summary-table, echo=FALSE}
lrt_display <- data.frame(
  Metric = c(
    'Genes in the aligned count/VST input',
    'Genes with maximum day-mean TMM CPM >= 10',
    'Genes passing LRT adjusted p-value < 1e-7',
    'Genes passing all filters and used for scoring'
  ),
  Value = c(
    format(lrt_summary$input_genes, big.mark = ','),
    format(lrt_summary$genes_passing_day_mean_tmm_cpm, big.mark = ','),
    format(lrt_summary$genes_passing_lrt_padj, big.mark = ','),
    format(lrt_summary$genes_passing_all_filters, big.mark = ',')
  ),
  stringsAsFactors = FALSE
)
knitr::kable(
  lrt_display,
  format = 'html',
  table.attr = 'class="sample-summary-table"',
  align = c('l', 'r'),
  caption = 'Sequential temporal-gene filtering.'
)
```

The heatmap displays the first 1,500 final temporal genes after ordering the LRT results by adjusted p-value and then raw p-value. PCA, scoring, and validation use the complete temporal set rather than this display subset.

```{r temporal-heatmap, echo=FALSE, results='asis'}
cat('<div id="fig-temporal-heatmap" class="figure-anchor"></div>')
tutorial_figure(
  'GSE122380_temporal_heatmap.png',
  'Heatmap of 1500 temporal genes across all samples, ordered by differentiation day.',
  '92%'
)
```

The four-cluster partition uses Ward-D2 hierarchical clustering of Euclidean distances between robust-LOESS-smoothed, re-standardized day-mean profiles. It is used only to interpret broad expression patterns. Cluster trajectories and GO Biological Process enrichment use the same 1,500 displayed genes, with those genes as the enrichment universe. These tests are not part of the scoring algorithm.

```{r temporal-clusters, echo=FALSE, results='asis'}
cat('<div id="fig-temporal-clusters" class="figure-anchor"></div>')
tutorial_figure(
  'GSE122380_temporal_clusters.png',
  'Four temporal expression clusters with their enriched GO biological processes.'
)
```

<div class="tutorial-figure-caption"><strong>Figure 3.</strong> Descriptive temporal clusters and GO enrichment for the 1,500 heatmap genes. Each row pairs a cluster trajectory with its top enriched GO Biological Process terms.</div>

## Train the PCA reference space

Let \(X \in \mathbb{R}^{n \times p}\) be normalized reference expression for \(n\) samples and \(p\) temporal genes. For each of \(m\) reference timepoints, the scorer first calculates its mean expression profile \(\mathbf{m}_j\). PCA is fitted to the \(m \times p\) matrix of timepoint means after centering each gene without variance scaling. If \(\boldsymbol{\mu}_M\) is the across-timepoint gene mean and \(W_k\) contains the retained loading vectors, then

\[
\mathbf{m}_j = \frac{1}{|I_j|}\sum_{i \in I_j}\mathbf{x}_i,
\qquad
\mathbf{z}_i = (\mathbf{x}_i - \boldsymbol{\mu}_M) W_k.
\tag{1}
\]

The scorer chooses the smallest \(k\) for which cumulative PCA variance is at least 99%. This choice is made from reference-timepoint means only; samples being evaluated cannot affect it. The rule retained seven PCs in the complete reference fit and in every validation fold. PC1 is flipped for figure readability when needed so it increases with differentiation day; a sign flip does not alter Euclidean projections or scores.

```{r pca-day, echo=FALSE, results='asis'}
cat('<div id="fig-pca-day" class="figure-anchor"></div>')
tutorial_figure(
  'GSE122380_pca_day.png',
  'PCA and PC1 versus day for all temporal genes and early and late cluster subsets.'
)
```

<div class="tutorial-figure-caption"><strong>Figure 4.</strong> Individual samples projected into centered, unscaled PCA spaces learned from timepoint means for the full temporal set and two descriptive cluster subsets. C1+C2 and C3+C4 are interpretive subsets, not alternative primary models.</div>

The loading analysis annotates the largest PC1 and PC2 loadings and tests the top 500 genes in each PC1 direction for GO enrichment against the full temporal-gene universe.

```{r pc1-validation, echo=FALSE, results='asis'}
cat('<div id="fig-pc1-validation" class="figure-anchor"></div>')
tutorial_figure(
  'GSE122380_pc1_validation.png',
  'Top PCA loading genes and GO enrichment for positive and negative PC1 loadings.'
)
```

<div class="tutorial-figure-caption"><strong>Figure 5.</strong> PCA loading interpretation. <strong>a-b.</strong> The ten largest positive and negative PC1 and PC2 loading genes. Ambiguous Ensembl-to-symbol mappings retain the Ensembl ID. <strong>c-d.</strong> GO enrichment among the 500 strongest negative and positive PC1 loadings.</div>

## Fit and score the day-centroid polyline

For each ordered reference day \(t_j\), the mean PCA coordinate is

\[
\mathbf{c}_j = \frac{1}{|I_j|}\sum_{i \in I_j}\mathbf{z}_i,
\qquad j = 1,\ldots,m.
\tag{2}
\]

The ordered centroids define a piecewise-linear trajectory. Segment \(j\) begins at \(\mathbf{a}_j=\mathbf{c}_j\), ends at \(\mathbf{b}_j=\mathbf{c}_{j+1}\), and has direction \(\mathbf{v}_j=\mathbf{b}_j-\mathbf{a}_j\). For sample \(i\), the fraction along segment \(j\) is

\[
\alpha_{ij} =
\operatorname{clip}_{[0,1]}
\left(
\frac{\langle \mathbf{z}_i-\mathbf{a}_j,\mathbf{v}_j \rangle}
{\langle \mathbf{v}_j,\mathbf{v}_j \rangle}
\right).
\tag{3}
\]

The closest finite-segment projection is selected:

\[
\mathbf{q}_{ij}=\mathbf{a}_j+\alpha_{ij}\mathbf{v}_j,
\qquad
j^\ast=\underset{j}{\operatorname{argmin}}
\left\lVert\mathbf{z}_i-\mathbf{q}_{ij}\right\rVert^2.
\tag{4}
\]

The corresponding predicted time and normalized score are

\[
\widehat{t}_i=t_{j^\ast}+\alpha_{ij^\ast}(t_{j^\ast+1}-t_{j^\ast}),
\qquad
s_i=\frac{\widehat{t}_i-t_{\mathrm{start}}}{t_{\mathrm{end}}-t_{\mathrm{start}}}.
\tag{5}
\]

Zero-length centroid segments are omitted. The method assumes that ordered centroids describe a meaningful trajectory in retained PC space. A self-approaching or folded trajectory can make nearest-segment assignments biologically ambiguous; squared projection distance should therefore be inspected when adapting the scorer.

```{r timing-score-code, eval=FALSE, echo=TRUE}
source('functions/score_differentiation_timing.R')

timing_fit <- score_differentiation_timing(
  expression_matrix = vst,
  metadata = metadata,
  temporal_genes = temporal_genes,
  sample_id_col = 'sample_id',
  time_col = 'day_numeric'
)
head(timing_fit$scores)
```

```{r timing-polyline, echo=FALSE, results='asis'}
cat('<div id="fig-timing-polyline" class="figure-anchor"></div>')
tutorial_figure(
  'GSE122380_timing_polyline.png',
  'Samples and the ordered day-centroid polyline in the first two principal components.',
  '86%'
)
```

<div class="tutorial-figure-caption"><strong>Figure 6.</strong> Differentiation trajectory in reference PCA space. Black points are day centroids; arrows connect them in day order. The day 1 and day 15 centroids anchor scores 0 and 1.</div>

The next figure projects the reference samples back onto their fitted trajectory. This is a descriptive calibration view, not out-of-sample validation. The gray band is the day-level mean score plus or minus one standard deviation.

```{r score-by-day, echo=FALSE, results='asis'}
cat('<div id="fig-score-by-day" class="figure-anchor"></div>')
tutorial_figure(
  'GSE122380_score_by_day.png',
  'Differentiation timing scores across observed differentiation days.'
)
```

<div class="tutorial-figure-caption"><strong>Figure 7.</strong> Reference-sample differentiation timing scores by observed day. Dashed lines mark the bounded reference endpoints at 0 and 1.</div>

# Internal cell-line cross-validation

Leave-one-cell-line-out validation repeats temporal-gene selection, PCA training, and polyline fitting using the remaining cell lines, then scores samples from the held-out cell line. The count-based DESeq2 LRT is refit in each fold. The tracked VST matrix is shared across folds because the upstream transformation pipeline is not available; the result is therefore internal validation conditional on this processed cohort, not an external transferability estimate. The tutorial uses a compatible optional local cache when present and otherwise runs the complete validation in memory without saving the result.

Fold-specific temporal genes are also divided into maturation and progenitor sets by the sign of their Spearman correlation with day. The reported accuracy metrics use the complete temporal set. All 13 held-out cell lines contribute; cell line `19190` is omitted only from the per-cell-line panel grid so the display remains a balanced 3-by-4 layout.

```{r loo-summary-table, echo=FALSE}
loo_display_summary <- loo_accuracy_summary
names(loo_display_summary) <- c('Metric', 'Value')
knitr::kable(
  loo_display_summary,
  format = 'html',
  table.attr = 'class="sample-summary-table"',
  align = c('l', 'r'),
  caption = 'Leave-one-cell-line-out validation summary.'
)
```

```{r loo-cell-line-predictions, echo=FALSE, results='asis'}
cat('<div id="fig-loo-cell-line-predictions" class="figure-anchor"></div>')
tutorial_figure(
  'GSE122380_loo_cell_line_predictions.png',
  'Predicted versus observed day trajectories for twelve displayed held-out cell lines.'
)
```

<div class="tutorial-figure-caption"><strong>Figure 8.</strong> Held-out predictions for 12 displayed cell lines. Each panel reports Pearson \(r\) and its square for that line; neither quantity is a predictive \(R^2\) statistic.</div>

```{r loo-summary-figure, echo=FALSE, results='asis'}
cat('<div id="fig-loo-summary" class="figure-anchor"></div>')
tutorial_figure(
  'GSE122380_loo_summary.png',
  'Predicted versus observed differentiation day for every held-out sample.'
)
```

<div class="tutorial-figure-caption"><strong>Figure 9.</strong> Predicted versus observed differentiation day for all held-out samples. The dashed line is identity; the solid line is an ordinary least-squares summary.</div>

For held-out sample \(i\), absolute prediction error is

\[
e_i=\left|\widehat{t}_i-t_i\right|.
\]

Bars show the mean absolute held-out prediction error at each observed day. Error bars show one standard deviation across held-out samples at that day.

```{r loo-timepoint-accuracy, echo=FALSE, results='asis'}
cat('<div id="fig-loo-timepoint-accuracy" class="figure-anchor"></div>')
tutorial_figure(
  'GSE122380_loo_timepoint_accuracy.png',
  'Mean absolute held-out prediction error by observed differentiation day, with standard-deviation error bars.',
  '82%'
)
```

<div class="tutorial-figure-caption"><strong>Figure 10.</strong> Absolute held-out prediction error by day. Lower values indicate closer timing predictions.</div>

# Interpretation and adaptation

The score estimates position along this reference trajectory. It does not establish cell quality, differentiation efficiency, lineage identity, potency, or biological directionality.

For a new study, select temporal genes and train the PCA/polyline model in an appropriate reference cohort, then project test samples without refitting that model. Before comparing scores across datasets, verify gene identifiers, normalization, shared feature support, batch structure, reference-time coverage, and projection distance.

# References

1. Love MI, Huber W, Anders S. Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2. *Genome Biology*. 2014;15:550.
2. Strober BJ, Elorbany R, Rhodes K, et al. Dynamic genetic regulation of gene expression during cellular differentiation. *Science*. 2019;364:1287-1290.

# Session information

```{r session-info}
session_info_lines <- sub(
  '[[:space:]]+$',
  '',
  utils::capture.output(utils::sessionInfo())
)
session_info_lines <- session_info_lines[nzchar(session_info_lines)]
cat(session_info_lines, sep = '\n')
```
